---
title: "Finding the most severe storms in the US"
author: "Juan C. López Tavera"
date: "9/22/2017"
output: html_document
---


```{r setup, include=TRUE, message=FALSE}
# Setting up the report
## Loading the necessary package for reproducing the assingment
if (!require(data.table)) {
    install.packages("data.table")
}
if (!require(R.utils)) {
    install.packages("R.utils")
}

if (!require(knitr)) {
    install.packages("knitr")
}
library(knitr)

if (!require(tidyverse))
    install.packages("tidyverse")
library(tidyverse)

if (!require(datasets)) {
    install.packages("datasets")
}
library(datasets)

## Setting up all code chunks according to the assignment specs
knitr::opts_chunk$set(
    eval = TRUE,
    echo = TRUE,
    tidy = TRUE,
    results = "markup",
    include = TRUE,
    message = FALSE,
    warning = FALSE,
    knitr.table.format = "markdown",
    tidy.opts = list(width.cutoff = 80),
    fig.align = "center",
    fig.path = "figure/",
    highlight = TRUE, 
    cache = TRUE
)

```


## Introduction

This is a reproducible research report made with knitr for completing the requirements of the JHU's Reproducible Research Course. 

This documents describe the exploration of U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database, which tracks weather phenomena, their characteristics and casualties caused by them. 

The main objective of this data exploration is to answer the questions:

1. Across the USA, which types of events are most harmful with respect to population health?
2. Across the USA, which types of events have the greatest economic consequences?


## Data downloading and reading

First, if necessary, we donwload the data BZ2 file from the link provided in the course assignment page; then, we read the data into the working environment, and conveniently cache the result of the computation.  

```{r data loading, cache=TRUE, message=FALSE}
URL <-
  "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
path <- paste0(getwd(), "/repdata%2Fdata%2FStormData.csv.bz2")
csvfile <- gsub(pattern = ".bz2",
                replacement = "",
                x = path)

if (!file.exists(path)) {
  download.file(url = URL,
                destfile = path)
  R.utils::bunzip2(path, remove = FALSE)
  storm_data <- data.table::fread(input = csvfile, 
                                  showProgress = FALSE, 
                                  verbose = FALSE)
  
} else if (!file.exists(csvfile)) {
  R.utils::bunzip2(path, remove = FALSE)
  storm_data <- data.table::fread(input = csvfile, 
                                  showProgress = FALSE, 
                                  verbose = FALSE)
  
} else {
  storm_data <- data.table::fread(input = csvfile, 
                                  showProgress = FALSE, 
                                  verbose = FALSE)
}

```


## Data Exploration

We need to know our dataset, so we explore its structure: 

```{r data structure, include=TRUE, eval=FALSE}
str(storm_data)

```

The output from the code above is very lengthy, given the number of variables in our dataset, so we don't include it for aesthetic purposes. 

We have a tibble data frame of `r nrow(storm_data)` rows by `r ncol(storm_data)` columns, with `r sum(is.na(storm_data))` missing values, which are `r round(mean(is.na(storm_data))*100, 2)`% of the observations.

The classes of the NOAA storm database, `storm_data`, and the proportion of missing values for each column are shown in the following table:

```{r data processing}

classes <-
sapply(storm_data, class) %>% 
        as_data_frame() %>% 
        rownames_to_column("Variable") %>% 
        select(Variable, Type = value)

classes$`Percent of NAs` <- sapply(storm_data, function(x)
        round(mean(is.na(x)) * 100, 2))

kable(classes %>% 
              arrange(desc(`Percent of NAs`)) %>%
              filter(`Percent of NAs` != 0)
              )

```

The percentage of missing values per variable goes from 0% of NAs to as much as 100% of NAs! Variables with 0% of NAs are ommited from the table above for aesthetic purposes. 




## Results

### Most harmful meteorological events

Across the United States, which types of events (as indicated in the `EVTYPE` variable) are most harmful with respect to population health?

```{r harmful events}

```


### Most costly meteorological events

Across the United States, which types of events have the greatest economic consequences?

```{r costly events}

```





